# Exercises

These are intended to be done **after** completing the worked examples.

## Exercise 1 — <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119732>

Using **GSE119732**, confirm whether the ID column contains Ensembl IDs with version suffixes.

1.  Extract the first 20 IDs.
2.  Count how many contain a `.`.
3.  Create a new column with versions stripped.
4.  Map the identifiers to HGNC symbols.

```{r}
source("./fetch_geo_supp.R")
library(dplyr)
library(readr)
library(biomaRt)

gse <- "GSE119732"
fetch_geo_supp(gse = gse)

path  <- file.path("data", gse)
files <- list.files(path, full.names = TRUE, recursive = TRUE)

# pick the first table-like file (adjust pattern if needed)
f <- files[grepl("\\.(tsv|txt|csv)(\\.gz)?$", files, ignore.case = TRUE)][1]
x <- readr::read_tsv(f, show_col_types = FALSE)

id_col <- names(x)[1]
ids20  <- x[[id_col]] |> head(20) |> as.character()

# 1) Extract first 20 IDs + 2) Count how many contain a "."
n_with_dot <- sum(grepl("\\.", ids20))

strip_ensembl_version <- function(x) {
  sub("\\..*$", "", x)
}

ids_preview <- tibble(
  raw_id    = ids20,
  has_dot   = grepl("\\.", ids20),
  stripped  = strip_ensembl_version(ids20)
)

ids_preview
n_with_dot

# 3) Create stripped column in the full table
x2 <- x %>% mutate(ensembl_gene_id = strip_ensembl_version(.data[[id_col]]))

# 4) Map to HGNC
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(x2$ensembl_gene_id),
  mart       = ensembl
)

expr_mapped <- x2 %>% left_join(map, by = "ensembl_gene_id")
expr_mapped %>% dplyr::select(ensembl_gene_id, hgnc_symbol, everything()) %>% head()
```

## Exercise 2 — <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122380>

Using **GSE122380**, confirm whether the ID column contains Ensembl IDs with version suffixes.

1.  Extract the first 20 IDs.
2.  Create a new column with versions stripped.
3.  Map the identifiers to HGNC symbols.
4.  What is different about this file?

```{r}
source("./fetch_geo_supp.R")
library(dplyr)
library(readr)
library(biomaRt)

strip_ensembl_version <- function(x) sub("\\..*$", "", x)

gse <- "GSE122380"
fetch_geo_supp(gse = gse)

path  <- file.path("data", gse)
files <- list.files(path, full.names = TRUE, recursive = TRUE)

f <- files[grepl("raw_counts|counts|\\.txt|\\.tsv|\\.csv", basename(files), ignore.case = TRUE)][1]
x <- safe_read(f)

id_col <- names(x)[1]
ids20  <- head(as.character(x[[id_col]]), 20)

# first 20 + stripped
ids_preview <- tibble(
  raw_id   = ids20,
  has_dot  = grepl("\\.", ids20),
  stripped = strip_ensembl_version(ids20)
)
ids_preview

# full table with stripped IDs
x2 <- x %>% mutate(ensembl_gene_id = strip_ensembl_version(.data[[id_col]]))

# map Ensembl -> HGNC
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
map <- getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters    = "ensembl_gene_id",
  values     = unique(x2$ensembl_gene_id),
  mart       = ensembl
)

expr_mapped <- x2 %>% left_join(map, by = "ensembl_gene_id")
expr_mapped %>% dplyr::select(ensembl_gene_id, hgnc_symbol, dplyr::everything()) %>% head()
```

On the GEO series page for GSE122380, the “processed data” includes two supplementary files: an Excel table and the raw counts matrix (`GSE122380_Supplementary_Data_Table_S1.xlsx` and `GSE122380_raw_counts.txt.gz`).

## Exercise 3

Can you use the worked example to process the above two GEO records? How?

```{r}
process_geo_counts <- function(gse) {
  fetch_geo_supp(gse = gse)
  path  <- file.path("data", gse)
  files <- list.files(path, full.names = TRUE, recursive = TRUE)

  f <- files[grepl("counts|raw|\\.txt|\\.tsv|\\.csv", basename(files), ignore.case = TRUE)][1]
  x <- safe_read(f)

  id_col <- names(x)[1]
  ids20  <- head(as.character(x[[id_col]]), 20)

  ids_preview <- tibble(
    raw_id   = ids20,
    has_dot  = grepl("\\.", ids20),
    stripped = strip_ensembl_version(ids20)
  )

  x2 <- x %>% mutate(ensembl_gene_id = strip_ensembl_version(.data[[id_col]]))

  ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
  map <- getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters    = "ensembl_gene_id",
    values     = unique(x2$ensembl_gene_id),
    mart       = ensembl
  )

  expr_mapped <- x2 %>% left_join(map, by = "ensembl_gene_id")
  list(file = f, ids_preview = ids_preview, mapped_head = head(expr_mapped))
}

# run both exercises
res_6_1 <- process_geo_counts("GSE119732")
res_6_2 <- process_geo_counts("GSE122380")

res_6_1$ids_preview
res_6_2$ids_preview
```

Yes: same pipeline, just parameterize the GSE and let the code:

-   fetch supp files

-   read the counts table

-   grab first 20 IDs + count dots

-   strip versions

-   map via biomaRt
