# Exercises

These are intended to be done **after** completing the worked examples.

## Exercise 1 — <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119732>

Using **GSE119732**,

1.  Map the distribution of the samples as a box plot and a density plot using only the raw data.

    ```{r}
    library(GEOquery)
    library(readr)
    library(edgeR)

    # Download + read
    getGEOSuppFiles("GSE119732", makeDirectory=TRUE, baseDir="data")
    files <- list.files("data/GSE119732", pattern="\\.gz$", full.names=TRUE, recursive=TRUE)
    x <- read_tsv(files[1], show_col_types=FALSE)

    counts <- as.matrix(x[,-1])
    rownames(counts) <- x[[1]]

    # RAW DISTRIBUTIONO (BOX AND DENSITY)
    boxplot(counts, las=2, main="Raw counts")

    plot(density(as.numeric(counts)), main="Raw counts density")
    ```

2.  Add colours to the plot separating the different variables in the model

    ```{r}
    samples <- colnames(counts)
    celltype <- substr(samples, 2, nchar(samples))
    cols <- as.numeric(as.factor(celltype))

    boxplot(counts, las=2, col=cols, main="Raw counts by celltype")

    plot(density(as.numeric(counts)), main="Raw density by celltype")
    legend("topright", legend=levels(as.factor(celltype)),
           col=1:length(unique(cols)), lty=1)
    ```

3.  Filter out lowly expressed genes and re plot the distributions.

    ```{r}
    dge <- DGEList(counts=counts)
    keep <- filterByExpr(dge)
    dge <- dge[keep,,keep.lib.sizes=FALSE]

    boxplot(dge$counts, las=2, col=cols,
            main="Filtered counts")

    plot(density(as.numeric(dge$counts)),
         main="Filtered density")
    ```

## Exercise 2 — <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122380>

Using **GSE122380**, confirm whether the ID column contains Ensembl IDs with version suffixes.

1.  Map the distribution of the samples as a box plot and a density plot using only the raw data.

    ```{r}
    getGEOSuppFiles("GSE122380", makeDirectory=TRUE, baseDir="data")
    files <- list.files("data/GSE122380", pattern="\\.gz$", full.names=TRUE, recursive=TRUE)
    x <- read_tsv(files[1], show_col_types=FALSE)

    counts <- as.matrix(x[,-1])
    rownames(counts) <- x[[1]]

    # CONFIRM ENSMEBL VERSION SUFFIX
    head(x[[1]], 20)
    sum(grepl("\\.", x[[1]][1:20]))

    # RAW DISTRIBUTION # comment back to get the plot
    # boxplot(counts, las=2, main="Raw counts")
    # plot(density(as.numeric(counts)), main="Raw counts density")
    ```

2.  Add colours to the plot separating the different variables in the model

    ```{r}
    samples <- colnames(counts)
    group <- substr(samples, 1, 1)
    cols <- as.numeric(as.factor(group))

    # boxplot(counts, las=2, col=cols, # comment back to get the plot
    #         main="Raw counts by group")
    ```

3.  Filter out lowly expressed genes and re plot the distributions.

    ```{r}
    dge <- DGEList(counts=counts)
    keep <- filterByExpr(dge)
    dge <- dge[keep,,keep.lib.sizes=FALSE]

    # boxplot(dge$counts, las=2, col=cols, # comment back to get the plot
    #         main="Filtered counts")
    ```

## Exercise 3 -

Can you use the worked example to process the above two GEO records? How?

Yes! You can do:

-   Raw counts

    -   DGEList()

    -   filterByExpr()

    -   calcNormFactors(method="TMM")

    -   cpm()

    -   plotMDS()
