# Exercises: Inspect Specific GEO Series

In this exercise, you'll retrieve metadata for the following GEO Series and answer:

**GSE IDs:** "GSE201427","GSE166530","GSE162183","GSE186458"

**Questions per GSE**

1.  **What type of data is it? Describe it.** (platform technology; e.g., high-throughput sequencing, array)
2.  **What disease or model is it using?** (infer from series title/summary/platform organism)
3.  **How many samples are there? What type of data is associated with the record?**
4.  **How many control and how many disease?**
5.  \*\*Can you find out any information about the GEO dataset - GSE264108. Why or Why not?

> Hint: Sample labels are not standardized. Use robust keyword detection (e.g., `control|normal|healthy|wild[- ]?type` vs. `tumou?r|cancer|patient|disease|mutant|treated`). Inspect a few GSM rows if the automatic inference is uncertain.

# Answers

### First, understand the data

```{r}
dbListTables(con)
```

```{r}
dbListFields(con, 'gse')
```

### Q1 + Q2

```{r}
sql <- paste(
  "SELECT DISTINCT",
  "gse.gse,",
  "gse.title AS gse_title,",
  "gse.summary,",
  "gse.overall_design,",
  "gse.type AS gse_type,",
  "gse.submission_date,",
  "gpl.gpl,",
  "gpl.title AS gpl_title,",
  "gpl.technology,",
  "gpl.organism",
  "FROM gse",
  "LEFT JOIN gse_gpl ON gse_gpl.gse = gse.gse",
  "LEFT JOIN gpl ON gpl.gpl = gse_gpl.gpl",
  "WHERE gse.gse = 'GSE264108'"
)

rs <- dbGetQuery(con, sql)
rs
```

### Q3

```{r}
sql <- paste(
  "SELECT",
  "gsm.gsm,",
  "gsm.title,",
  "gsm.source_name_ch1,",
  "gsm.characteristics_ch1,",
  "gsm.description,",
  "gsm.organism_ch1",
  "FROM gse_gsm",
  "JOIN gsm ON gsm.gsm = gse_gsm.gsm",
  "WHERE gse_gsm.gse = 'GSE264108'"
)

gsm_rs <- dbGetQuery(con, sql)
gsm_rs

paste("There are", nrow(gsm_rs), "samples.")
```

### Q4

```{r}
library(dplyr)
library(stringr)

gsm_labeled <- gsm_rs %>%
  mutate(
    blob = str_to_lower(
      paste(title, source_name_ch1, characteristics_ch1, description, sep=" ")
    ),
    group = case_when(
      str_detect(blob, "control|normal|healthy|wild[- ]?type") ~ "control",
      str_detect(blob, "tumou?r|cancer|disease|patient|case|mutant|treated") ~ "disease",
      TRUE ~ "unknown"
    )
  )

gsm_labeled %>% count(group)

# Over here, we may not be able to get clear groups... if that's the case, it's best to just inspect the rows from gsm_rs manually
```
